Are we learning to build better models at the cost of better decisions?

### 2021-03-11 manuscript/appendix_B.Rmd
appendix_B and notes

  d %>% pivot_longer(names(model_set[-1]), "model") %>%
  ggplot(aes(state, value, col=as.numeric(model))) +
  geom_hline(aes(yintercept = 0), lwd=1) + 
  geom_point(alpha = .5) + 
  geom_line(aes(state, true), col = "red") +
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
```

@@

am1 <- adaptive_management(transitions[-1], 
                           reward, 
                           discount, 
                           x0 = x0, 
                           Tmax = 50, 
                           true_transition = transitions[[true_model]], 
                           epsilon = 0.001, 
                           max_iter = 2000)
```

```{r}
am <- am1$df %>% mutate(belief = am1$posterior[,1])
am %>%  ggplot(aes(time, states[state], col = belief)) + 
  geom_line() + 
  geom_point() +
  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
```

```{r}

am_npv <- sum(am$value * discount ^ am$time)
am_npv / npv[[1,"npv"]]
```

>>>>

d %>% pivot_longer(names(model_set[-1]), "model") %>%
  mutate(model = as.numeric(model)) %>%
  ggplot(aes(state, value, col = model)) +
  geom_hline(aes(yintercept = 0), lwd=1) + 
  geom_point(alpha = .5) + 
  geom_line(aes(state, true), col = "red") +
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
```

@@

am_many <- adaptive_management(transitions[-1], 
                           reward, 
                           discount, 
                           x0 = x0, 
                           Tmax = 50, 
                           true_transition = transitions[[true_model]], 
                           epsilon = 0.001, 
                           max_iter = 2000)
```

```{r}
i <- 1:5
posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior[i,])) %>%
  pivot_longer(starts_with("X"), 
               values_to = "probability",
               names_to = "time") %>% 
  mutate(time = as.integer(as.factor(time))) %>%
  pivot_longer(c("r", "K"), names_to="parameter")

posteriors %>%
  ggplot(aes(value, probability, group=time, col=time)) +
  geom_line(show.legend = FALSE) +
  facet_grid(~parameter, scales="free")

```{r, fig.cap = ""}
am_many$df %>% 
  mutate(belief = am_many$posterior[,1]) %>%
  ggplot(aes(time, states[state], col = belief)) + 
  geom_line() + 
  geom_point() +
  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) + 
  xlab("fish stock")
```

```{r}
am_npv <- sum(am$value * discount ^ am$time)
am_npv / npv[[1,"npv"]]
```

### 2021-03-11 manuscript/manuscript.Rmd
appendix_B and notes

<!-- 
First: Intuition for the result.
  - See that model 1 captures the key feature, while model 2 is just 'closer to the curve'
  - See that this is a generic truth about decision-making: there's always a Model 1
  - This is not about over-fitting.

-->
Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other. Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.  In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values. This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Appendix B, Fig S1).

The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.  Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.  While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.  This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.

This phenomenon is not unique to sequential decision problems or optimal control solutions.  For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.  This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.  Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem. Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2. Does this mean MSE is not susceptible to this issue?  No, it does not.  It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.  Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954]. If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.

Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [].  In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.  The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy. But in the example here, both Model 1 and Model 2 have the same structural complexity.  The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.

This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2. Model 1 approximates the key feature, giving nearly optimal outcomes.  Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood []. Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations. In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).  Models will always be simpler than reality.  To insist that this issue can and will be avoided by always including more and better models of the process misses the point.

<!--
So what?

- This is different than previous work: e.g. VOI is not zero, it's negative here.
- Model 1 is not a unicorn: we cannot rule out these situations in real world (particularly ecology where we don't actually do much iterative updating in practice)
- 
-->

<!-- **Is it just a unicorn?** -->
Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?  It is tempting to argue that Model 1 is a kind of unicorn, a mythical creature existing only in theory. But that is a difficult assertion to prove.  Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [@].  If there is no shortage of bad forecasts, then could any of them be useful?  The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.  Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].  Capturing only the essential aspects as simply as possible is the goal of any model building exercise.  Thus it should be no surprise that a model can drive good decisions while making poor predictions.  It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management or iterative forecasting [@Dietze2018].  This makes it more likely that such models could persist in practice until now. While iterations that revise this models will no doubt improve decision outcomes in many case, it is worth bearing in mind from the example here that such a connection is not guaranteed.

<!-- So what? -->
If these unicorns do exist in real world management, then what do we do about them? I believe this is an open question, but that our first step must be to recognize it as such.  (1) that the issue is just an example of the already-understood problem of underspecifying model uncertainty, which is resolved by including more or better models, or (2) this issue is just an example of the already-understood limitations of relying too much on models to shape decision-making. Either we need more models, or we need fewer models.

If to advocates of iterative model updating approaches the solution is always more models, then to skeptics the solution is fewer models and algorithms.  
If we cannot always resolve the issue by adding more models, is this just another example of the risks of an over-reliance on models at all?

```{r figure4,  fig.cap="Panel A: Population growth curves of each model. The positive equilibrium of each model occurs where the curve crosses the horizontal axis. Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.", fig.width=8}
# Fig 4
model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
policies <- read_csv("../data/policies.csv") %>% mutate(model = as.character(model))

@@

<!-- move to appendix?  too confusing/too long?

Not sure - it feels a little extraneous (I did think this was an interesting point brought up by reviewer 1 but I also don't necessarily think it warrants a long paragraph that might confuse the main point rather than clarify something) Could this be a couple sentences and reference an appendix that walks through what you write out (potentially even with a simple constant mortality example)
-->
Is this result an artifact of optimal control? The analysis has relied on optimal control theory to determine the optimal decision associated with each model or distribution of models.  The resulting policies are mathematically optimal under the assumptions given [see @Reed1979 for stochastic harvest problem in particular and @Marescot2013 for discussion of Markov Decision Problems more generally].  In most realistic decision problems, model complexity currently precludes the calculation of an optimal control solution at all ["the curse of dimensionality", e.g. @Polasky2011; @Marescot2013]. Under the assumptions of any of the models considered here, the optimal control solution leads to what is known as a "constant escapement" policy, which seeks to bring biomass to the target escapement level, $B_{MSY}$ as quickly as possible, $H_t = \min(X_t - B_{MSY}, 0)$.  If the assumptions are not met, a constant-escapement policy may not be optimal, as in the case when observations of the stock $X_t$ are not perfect [@pomdp-intro].  While constant escapement is used as a basis of management in some real world cases such as salmon fisheries, most marine fisheries are based on a target of "constant mortality", ($F_{MSY}$). Notably, a constant mortality policy, implemented as $H_t = F_{MSY} X_t$, would not perform better under Model 1 than Model 2.  Constant mortality is optimal only under deterministic growth at equilibrium (where the policy is the same under constant escapement).  Does that mean the result is an artifact of assuming constant escapement rather than constant mortality as the basis for management?  No, it does not.  It is just as easy to construct a model under which the constant mortality matches the constant mortality of the true model while being a poor predictor.  Moreover, using a non-optimal policy would always beg the question of whether the poor outcomes were merely the result of using a non-optimal policy rather than a consequence of model uncertainty.  Because the space of possible actions is smaller than that of possible models, we should expect most decision problems to have their own "Model 1," a model that would lead to correct actions while being a poor predictor of future outcomes.  The trouble is that we have no theory which would not reject such a model.

Is this result merely an artifact of differing objectives?  @Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.  However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.  Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1. To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.  Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].

<!-- I really like the last paragraph!! -->

Perhaps Model 1 is a unicorn, a mythically simple model that gives optimal results but has no analogue in the real world. This is not an easy assertion to prove either way.  Future observations consistently fall outside the 95% confidence intervals predicted under Model 1.  Unfortunately, such innaccurate predictions are only too common in the real world.  Have any of those led to effective policy despite the innaccuracy?  Have any been replaced with more accurate models that have also driven worse decisions? That is more difficult to answer.  What this example demonstrates is that if such unicorn models exist, our current paradigm will surely filter them out eventually, to be replaced with worse outcomes.

<!-- Omit this?  Does it beg an example?

There is at least one technique which might address the problem presented here.  Recently developed techniques based on model-free deep reinforcement learning train multi-layer neural networks to approximate the optimal control policy without assuming or attempting to model the underlying dynamic process at all.  Feedback is based on cumulative reward, not short-term reward or predictive accuracy.  
-->

<!--I was thinking about exactly this while I was reading the paper.  I do think this would likely beg an example, which might be a can of worms that you don't want to open with this piece... I also don't think that I think it would address the presented problem. (Kind of repeating my second framing point but) It reminds me of the "Stop Explaining Black Box Machine Learning Models for High Stakes Decisions and use Interpretable Models Instead" paper where she talks about how, for a given data set, you can define the Rashomon set as the set of reasonably accurate predictive models-  Because the data are finite, the data could admit many close-to-optimal models that predict differently from each other: a large Rashomon set. I suspect this happens often in practice because sometimes many different ML algorithms perform similarly on the same data set, despite having different functional forms (for example, random forests, neural networks, support vector machines). As long as the Rashomon set contains a large enough set of models with diverse predictions, it probably contains functions that can be approximated well by simpler functions, and so the Rashomon set can also contain these simpler functions. Said another way, uncertainty arising from the data leads to a Rashomon set; a larger Rashomon set probably contains interpretable models, thus interpretable accurate models often exist. If this theory holds, we should expect to see interpretable models exist across domains. These interpretable models may be hard to find through optimization, but at least there is a reason we might expect that such models exist.

The problem is deeper than acknowleding a "rashamon set of models" or a plurality of reality (relevant to that Bioscience paper I sent last week about the Rashamon effect in conservation). Sometimes what appears closer to reality is actually fundementally wrong in some way that is misguiding to a decision maker...more opaque approaches could lessen our capacity to trouble shoot that?) -->

>>>>

<!-- 
First: Intuition for the result.
  - See that model 1 captures the key feature, while model 2 is just 'closer to the curve'
  - See that this is a generic truth about decision-making: there's always a Model 1
  - This is not about over-fitting.
-->

<!-- quick summary of what was demonstrated -->
Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other. Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.  In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values. This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Appendix B, Fig S1).

<!-- Intuition, Figure 4 -->
The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.  Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.  While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.  This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.

<!-- Inttuition for it's not just this example -->
This phenomenon is not unique to sequential decision problems or optimal control solutions.  For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.  This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.  Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem. Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2. Does this mean MSE is not susceptible to this issue?  No, it does not.  It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.  Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954]. If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.

```{r figure4,  fig.cap="Panel A: Population growth curves of each model. The positive equilibrium of each model occurs where the curve crosses the horizontal axis. Note that while Model 2 is a better approximation to the truth (Model 3), Model 1 better approximates the stock size which leads to maximum growth.  Panel B: The optimal control policy under Model 1 is nearly identical to that under the true Model 3, while the optimal policy under Model 2 supresses stock to a much lower escapement level.", fig.width=8}
# Fig 4
model_curves <- read_csv("../data/model_curves.csv") %>% mutate(model = as.character(model))
policies <- read_csv("../data/policies.csv") %>% mutate(model = as.character(model))

@@

<!-- Not just overfitting -->
Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@overfit].  In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.  The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy. But in the example here, both Model 1 and Model 2 have the same structural complexity.  The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.

<!-- Not just "get more models" -->
This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2. Model 1 approximates the key feature, giving nearly optimal outcomes.  Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood []. Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations. In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).  Models will always be simpler than reality.  To insist that this issue can and will be avoided by always including more and better models of the process misses the point.

<!--
So what?
- This is different than previous work: e.g. VOI is not zero, it's negative here.
- Model 1 is not a unicorn: we cannot rule out these situations in real world (particularly ecology where we don't actually do much iterative updating in practice)
- 
-->

<!-- **Is it just a unicorn?** -->
Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?  It is tempting to argue that Model 1 is a kind of unicorn, a mythical creature existing only in theory. But that is a difficult assertion to prove.  Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [@].  If there is no shortage of bad forecasts, then could any of them be useful?  The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.  Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].  Capturing only the essential aspects as simply as possible is the goal of any model building exercise.  Thus it should be no surprise that a model can drive good decisions while making poor predictions.  It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management [@Walters1978] or iterative forecasting [@Dietze2018].  This makes it more likely that such models could persist in practice until now. While iterations that revise this models will no doubt improve decision outcomes in many case, it is worth bearing in mind from the example here that such a connection is not guaranteed.

<!-- So what? -->
If these unicorns do exist in real world management, then what do we do about them? I believe this is an open question, but that our first step must be to recognize it as such.  We have seen that the problem cannot be resolved by more data, and is not the result of overfitting. Nor is "creating more models" the answer: when we have a model that is good enough to get optimal results, we cannot always insist on more models. We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations. Because value of information can be negative, we must revise model uncertainty only cautiously.  More importantly, we have seen that decision theory, such as the intuition offered by Fig 4, can help us better understand what features of a model are essential to decision outcomes and what are not.  Building on such understanding, it may possible to identify strategies for learning that are more agnostic to the details of the model that are not important.

<!-- Model free management -->

<!--
Is this result merely an artifact of differing objectives?  @Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.  However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.  Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1. To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.  Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].
-->
